### 2024-01-22 doc/article/article.tex
article
revision

\usepackage{mathtools}
\usepackage{tuda-pgfplots}
\usepackage{siunitx}
\usepackage{subcaption}

@@

@@

\abstract[Summary]{Electrical circuits are present in a variety of technologies, making their design an important part of computer aided engineering. The growing number of tunable parameters that affect the final design leads to a need for new approaches of quantifying their impact. Machine learning may play a key role in this regard, however current approaches often make suboptimal use of existing knowledge about the system at hand. In terms of circuits, their description via modified nodal analysis is well-understood. This particular formulation leads to systems of differential-algebraic equations (DAEs) which bring with them a number of peculiarities, e.g. hidden constraints that the solution needs to fulfill. We aim to use the recently introduced dissection concept for DAEs that can decouple a given system into ordinary differential equations, only depending on differential variables, and purely algebraic equations that describe the relations between differential and algebraic variables. The idea then is to only learn the differential variables and reconstruct the algebraic ones using the relations from the decoupling. This approach guarantees that the algebraic constraints are fulfilled up to the accuracy of the nonlinear system solver, which represents the main benefit highlighted in this article.}

\keywords{electrical circuits, differential-algebraic equations, modified nodal analysis, dissection concept, machine learning}

@@

\section{Introduction}
\label{sec:int}
Design optimization and uncertainty quantification are ubiquitous tools of modern computer aided  engineering. Due to the ever-increasing complexity of engineering systems, machine learning techniques have gained popularity for constructing surrogate models when the objective function becomes expensive to evaluate and a large number of (design or uncertainty) parameters are present. In these cases, classical model order reduction techniques\cite{schilders2008} or function approximation approaches\cite{xiu2010} suffer from the curse of dimensionality: the number of operations to construct and the memory required to store the model grow exponentially with respect to the number of parameters. Experimental and in some cases even theoretical evidence\cite{beneventano2020} shows that machine learning approaches may be able to overcome the curse of dimensionality and provide surrogates that are fast to evaluate, while requiring comparably little data.

In the particular case of electrical circuit design, neural networks have been used for design optimization for over 20 years\cite{zaabab1995,fang2000}. More recently, Gaussian process regression has been employed for both uncertainty quantification and design optimization during analog integrated circuit design\cite{wang2018,sanabria2020}. The commonality between these approaches is that they focus on the learning part: they aim to provide a computationally efficient and accurate surrogate, given data produced by some circuit simulation software. Thus, they all treat the circuit simulator as a black box that simply provides the data which is then used for constructing the surrogate. In contrast, we want to exploit the known structure that underlies the equations describing electrical circuits.

More specifically, we consider the modified nodal analysis\cite{ho1975} (MNA). MNA is one of the most popular circuit descriptions and lies at the center of SPICE-like simulation software such as LTspice\cite{ltspice}, Xyce\cite{xyce} and PSpice\cite{pspice}. Applying MNA to a given circuit leads to systems of differential-algebraic equations (DAEs), which can generally be written as systems of implicit differential equations \cite{denk1991}
\begin{align*}
    \mb{f}(\mb{x}', \mb{x}, t, \mb{p}) = \mb{0}, \quad \mb{x}(0) = \mb{x}_0,
\end{align*}
where the Jacobian $\mb{J}_{\mb{x}'}(\mb{f})$, of $\mb{f}$ w.r.t.~$\mb{x}'$, is singular and $\mb{p}$ are the design or uncertainty parameters. To give a more intuitive description, one can think of DAEs as ordinary differential equations (ODEs) that are constrained to a manifold determined by (hidden) constraints on the solution variables. Multiple so-called index concepts were introduced in order to characterize the difficulties that appear in the context of DAEs. A reference by Mehrmann\cite{mehrmann2015} outlines some of the most prominent, including the differentiation, tractability, strangeness and perturbation indices. For all of these, the index $\nu \in \mathbb{N}$ depends on structural properties of the DAE and (informally) measures the degree of difficulty associated with solving it. In this sense, higher index DAEs usually lead to more problems, both analytically and numerically.

Various relationships between the indices, including situations when they coincide, are also known\cite{mehrmann2015} and the structure and index of MNA are well understood for both independent sources\cite{tischendorf1999} and controlled sources\cite{estevez2000ijcta}. Rather recently, the dissection concept was introduced\cite{jansen2014}. It combines features of the strangeness and tractability indices to decouple a DAE into an ODE and a set of purely algebraic equations. In particular, the entire dynamics of the solution may then be found only using the ODE. We aim to use this dissection concept to propose an approach for learning electrical circuits more accurately and efficiently.

In the following, \autoref{sec:mna} introduces MNA in more detail and states some well-known related results. Afterwards, \autoref{sec:dc} outlines the dissection concept and showcases its properties using small example circuits. The same examples are then used to present the new approach on an abstract level in \autoref{sec:ial} and on a numerical level in \autoref{sec:ne}, where we also showcase an alternative method that trades some of the accuracy of our approach for an easier implementation. Preliminary conclusions about the effectiveness of our approach and future research directions are given in \autoref{sec:cfr}.

\section{Modified nodal analysis}
\label{sec:mna}
We begin by looking at the system of DAEs that results from MNA. From here onward we will disregard controlled sources. As these are of crucial importance for engineering applications however, we will state whether extensions including controlled sources are available or missing at the appropriate times. Borrowing the notation from Tischendorf\cite{tischendorf1999}, and only explicitly expressing the time dependence of independent sources for brevity, the system of MNA reads
\begin{subequations}
    \label{eq:mnai}
    \begin{align}
            \A{C} \ddt \qC \big( \AT{C} \vphi \big) + \A{R} \gR \big( \AT{R} \vphi \big) + \A{L} \i{L} + \A{V} \i{V} + \A{I} \i{s}(t) &= \mb{0}\\ 
            \ddt \phiL \big( \i{L} \big) - \AT{L} \vphi &= \mb{0}\\
            \AT{V} \vphi - \v{s}(t)&= \mb{0},
    \end{align}
\end{subequations}
where $\gR$, $\qC$ and $\phiL$ represent functions modeling resistive, capacitive or inductive devices respectively. The terms for independent current and voltage sources are given by $\i{s}(t)$ and $\v{s}(t)$, while $\vphi$ denotes the vector of nodal potentials and $\i{L} $, $\i{V}$ are the currents flowing through branches containing inductors or voltage sources respectively. The last ingredient is given by the incidence matrices $\A{*}$, where $*$ indicates the device type. These collect the branch to node relations of the underlying electrical network when considering the branches and nodes as edges and vertices of a graph. In the context of design optimization or uncertainty quantification, the device functions $\gR$, $\qC$ or $\phiL$ could e.g.~depend on (subsets of) the parameters $\mb{p}$.

In order to obtain a version of \eqref{eq:mnai} that is better suited to analysis and implementation, we consider the device function Jacobians
\begin{align}
    \mb{L}(\i{L}) \coloneqq \mb{J}_{\i{L}}(\phiL), \quad \mb{C} \big( \AT{C} \vphi \big) \coloneqq \mb{J}_{\AT{C} \vphi}(\qC) \label{eq:dfj},
\end{align}
where we use the same notation for the Jacobians as in \autoref{sec:int}. We also assume that the device function for resistors may be expressed as
\begin{align}
    \gR \big( \AT{R} \vphi \big) = \mb{G} \big( \AT{R} \vphi \big) \AT{R} \vphi \label{eq:dfr}.
\end{align}
In practice, this is accomplished by either considering the chordal (static) conductance $\frac{[\gR]_i}{[\AT{R} \vphi]_i}$ for each resistor $\mr{R}_i$ separately or by using the Jacobian $\mathbf{J}_{ \AT{R} \vphi}(\gR)$. Inserting \eqref{eq:dfj} and \eqref{eq:dfr} into the original system and writing it in matrix form yields
\begin{align}
    \begin{bmatrix}
        \A{C}^{\phantom{\T}} \mb{C} \big( \AT{C} \vphi \big) \AT{C} & &\\
        & \mb{L}(\i{L}) &\\
        & & \mb{0}
    \end{bmatrix} \ddt \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{R}^{\phantom{\T}} \mb{G} \big( \AT{R} \vphi \big) \AT{R} & \A{L} & \A{V}\\
        -\AT{L} & &\\
        -\AT{V} & &
    \end{bmatrix} \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{I} \i{s}(t)\\
        \mb{0}\\
        \v{s}(t)
    \end{bmatrix} = \mb{0} \label{eq:mnaii},
\end{align}
where we assume that $\qC$ and $\phiL$ do not explicitly depend on time. MNA is usually presented in the form of \eqref{eq:mnaii} and for it, there exists the following well-known topological index result.
\begin{quote}
    Assuming the matrix-valued functions $\mathbf{G}$, $\mathbf{L}$, $\mathbf{C}$ from \eqref{eq:dfj} and \eqref{eq:dfr} positive definite:
    \begin{enumerate}
        \item The index of \eqref{eq:mnaii} is $\nu \leq 2$.
        \item The index of \eqref{eq:mnaii} is $\nu \leq 1$, if and only if there are no loops consisting only of capacitors and voltage sources containing at least one voltage source and no cutsets consisting only of inductors and current sources.
    \end{enumerate}
    The result can be found in terms of the tractability index\cite{tischendorf1999}, the differentiation index \cite{estevez2000ijcta} and the dissection concept\cite{jansen2014}. % \cite[Theorem 4.28]{jansen2014}
\end{quote}
We remark that there also exist extensive results about when the index of MNA including controlled sources does not exceed $\nu = 2$\cite{estevez2000ijcta}.

\section{Dissection concept}
\label{sec:dc}
This section serves as a general introduction to the dissection concept and more specifically its application to MNA as given in \eqref{eq:mnaii}. Consider a linear DAE in standard form\cite{jansen2014},
\begin{align}
    \mb{M} \ddt \mb{x} + \mb{K} \mb{x} + \mb{f}(t) = \mb{0}, \quad \ker \mb{M} \supsetneq \{ \mb{0} \} \label{eq:dae},
\end{align}
where we again keep the time dependence of $\mb{x}$ implicit. We now demonstrate the first two steps of the dissection concept when applied to systems of the form of \eqref{eq:dae} and remark that although \eqref{eq:mnaii} is a nonlinear DAE, the decoupling still takes the same form in this case due to the special structure of MNA\cite{jansen2014}.

\subsection{Index one case}
Assuming $\mb{M} \in \mathbb{R}^{n \times m}$, we define four basis functions $\mb{P}$, $\mb{Q}$, $\mb{V}$, $\mb{W}$ such that
\begin{align*}
    \mr{im\, } \mb{Q} = \ker \mb{M}, \quad \mr{im\, } \mb{W} = \ker \mb{M}^{\T}
\end{align*}
and the columns of $\mb{P}$ and $\mb{Q}$ together form a basis of $\mathbb{R}^n$, while the columns of $\mb{V}$ and $\mb{W}$ together form a basis of $\mathbb{R}^m$.

\paragraph{Splitting the solution}
The key idea is to use the basis functions to split the solution $\mb{x}$ into two parts
\begin{align}
    \mb{x} = \mb{P} \mbt{x} + \mb{Q} \mbb{x} \label{eq:vs},
\end{align}
where $\tilde{\cdot}$ is used to indicate differential (dynamic) variables and $\bar{\cdot}$ signifies algebraic (fixed) variables. When inserting the splitting \eqref{eq:vs} into the general system \eqref{eq:dae} we observe the following
\begin{align}
    \mb{M} \mb{P} \ddt \mbt{x} + \mb{K} \mb{P} \mbt{x} + \mb{K} \mb{Q} \mbb{x} + \mb{f}(t) = \mb{0} \label{eq:rs}
\end{align}
and this motivates the notation, as only $\mbt{x}$ appears differentiated in time.

\paragraph{Splitting the system}
The procedure then continues by multiplying \eqref{eq:rs} with $\mb{V}^{\T}$ and $\mb{W}^{\T}$ from the left. This yields
\begin{subequations}
    \label{eq:io}
    \begin{align}
            \underbrace{\mb{V}^{\T} \mb{M} \mb{P}}_{\eqqcolon \mbt{M}} \ddt \mbt{x} + \underbrace{\mb{V}^{\T} \mb{K} \mb{P}}_{\eqqcolon \mbt{K}_\mb{P}} \mbt{x} + \underbrace{\mb{V}^{\T} \mb{K} \mb{Q}}_{\eqqcolon \mbt{K}_\mb{Q}} \mbb{x} + \underbrace{\mb{V}^{\T} \mb{f}(t)}_{\eqqcolon \mbt{f}(t)} &= \mb{0} \label{eq:ioa}\\
            \underbrace{\mb{W}^{\T} \mb{K} \mb{P}}_{\eqqcolon \mbb{K}_\mb{P}} \mbt{x} + \underbrace{\mb{W}^{\T} \mb{K} \mb{Q}}_{\eqqcolon \mbb{K}_\mb{Q}} \mbb{x} + \underbrace{\mb{W}^{\T} \mb{f}(t)}_{\eqqcolon \mbb{f}(t)} &= \mb{0} \label{eq:iob},
    \end{align}
\end{subequations}
where we also introduce shorthands for the arising products. The key observations regarding \eqref{eq:io} are the following: $\mbt{M}$ is regular by construction, thus if $\mbb{K}_\mb{Q}$ is regular, \eqref{eq:iob} gives a purely algebraic equation for $\mbb{x}$ in terms of $\mbt{x}$. Solving this equation for $\mbb{x}$ and inserting the solution into \eqref{eq:ioa} then yields an ODE in the differential variables $\mbt{x}$. In this case, the \textbf{(dissection) index} of the DAE is defined to be $\nu = 1$.

\subsection{Index two case}
There are of course many DAEs, including those described by MNA, which have a higher index. For these the dissection concept proceeds by introducing additional basis functions and continuing the splitting process in a similar fashion.

\paragraph{Splitting the algebraic variables}
We begin by considering basis functions $\mbb{P}, \mbb{Q}, \mbb{V}, \mbb{W}$ of $\mbb{K}_\mb{Q}$ and split the algebraic variables $\mbb{x}$ further as follows
\begin{align}
    \mbb{x} = \mbb{P} \mbb{x}_\mb{P} + \mbb{Q} \mbb{x}_\mb{Q} \label{eq:xb}.
\end{align}
Inserting this decomposition into \eqref{eq:iob} and multiplying once by $\mbb{V}^{\T}$ and $\mbb{W}^{\T}$ from the left yields
\begin{subequations}
    \label{eq:iti}
    \begin{align}
        \mbb{V}^{\T} \mbb{K}_\mb{P} \mbt{x} + \mbb{V}^{\T} \mbb{K}_\mb{Q} \mbb{P} \mbb{x}_\mb{P} + \mbb{V}^{\T} \mbb{f}(t) &= \mb{0} \label{eq:itia}\\
        \mbb{W}^{\T} \mbb{K}_\mb{P} \mbt{x} + \mbb{W}^{\T} \mbb{f}(t) &= \mb{0} \label{eq:itib}.
    \end{align}
\end{subequations}

\paragraph{Splitting the differential variables}
We now need basis functions $\mbt{P}, \mbt{Q}$ of $\mbb{W}^{\T} \mbb{K}_\mb{P}$ to split the differential variables $\mbt{x}$ as
\begin{align}
    \mbt{x} = \mbt{P} \mbt{x}_\mb{P} +
    \mbt{Q} \mbt{x}_\mb{Q} \label{eq:xt}
\end{align}
and inserting this splitting into \eqref{eq:itib} then allows for an explicit description of $\mbt{x}_\mb{P}$
\begin{align*}
    \mbt{x}_\mb{P} = -\big( \mbb{W}^{\T} \mbb{K}_\mb{P} \mbt{P} \big)^{-1} \mbb{W}^{\T} \mbb{f}(t) \eqqcolon \mbt{f}_\mb{P}(t),
\end{align*}
since $\mbb{W}^{\T} \mbb{K}_\mb{P} \mbt{P}$ is assumed to be regular (otherwise the DAE is underdetermined\cite{jansen2014}).

>>>>

\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{tuda-pgfplots}
\usepackage{siunitx}
\usepackage[european]{circuitikz}

@@

% \newtheorem{assumption}{Assumption}
% \newtheorem{theorem}{Theorem}
% \newtheorem{remark}{Remark}

\newcommand{\assumptionautorefname}{Assumption}
\renewcommand{\theoremautorefname}{Theorem}
\newcommand{\remarkautorefname}{Remark}

\usetikzlibrary{external}
\tikzexternalize

@@

\abstract[Summary]{
    Electrical circuits are present in a variety of technologies, making their design an important part of computer aided engineering. The growing number of parameters that affect the final design leads to a need for new approaches to quantify their impact. Machine learning may play a key role in this regard, however current approaches often make suboptimal use of existing knowledge about the system at hand. In terms of circuits, their description via modified nodal analysis is well-understood. This particular formulation leads to systems of differential-algebraic equations (DAEs) which bring with them a number of peculiarities, e.g. hidden constraints that the solution needs to fulfill. We use the recently introduced dissection index that can decouple a given system of DAEs into ordinary differential equations, only depending on differential variables, and purely algebraic equations, that describe the relations between differential and algebraic variables. The idea is to then only learn the differential variables and reconstruct the algebraic ones using the relations from the decoupling. This approach guarantees that the algebraic constraints are fulfilled up to the accuracy of the nonlinear system solver, and it may also reduce the learning effort as only the differential variables need to be learned.}

\keywords{electrical circuits, differential-algebraic equations, modified nodal analysis, dissection index, machine learning}

@@

\section{Introduction}
\label{sec:int}
Design optimization and uncertainty quantification are key tools of modern computer aided engineering, that both rely on objective functions to express quantities of interest in terms of the variables of the underlying system. Due to the increasing complexity of engineering systems, machine learning approaches have gained popularity for constructing surrogate models of objective functions when they become expensive to evaluate and a large number of (design or uncertainty) parameters are present. In such situations, classical model order reduction techniques\cite{schilders2008} or function approximation approaches\cite{xiu2010} suffer from the curse of dimensionality: the number of operations to construct and the memory required to store the surrogate model grow exponentially with respect to the number of parameters. Experimental and in some cases even theoretical evidence\cite{beneventano2020} shows that machine learning approaches may be able to overcome this curse of dimensionality and provide surrogate models that are fast to evaluate, while requiring comparably little data for their construction and storage.

In the context of electrical circuit design, neural networks have been used for design optimization for over 20 years\cite{zaabab1995,fang2000}. More recently, Gaussian process regression has been employed for both uncertainty quantification and design optimization during analog integrated circuit design\cite{wang2018,sanabria2020}. The commonality between these approaches is that they focus on the learning part: they all aim to provide a computationally efficient and accurate surrogate model, given data produced by some circuit simulator. Thus, they all treat the circuit simulator as a black box that simply provides the data which is then used for constructing the surrogate. In contrast, we want to exploit the known structure that underlies the equations describing electrical circuits.

More specifically, we consider the modified nodal analysis\cite{ho1975} (MNA). MNA is one of the most popular circuit descriptions and lies at the center of SPICE-like simulation software such as LTspice\cite{ltspice}, Xyce\cite{xyce} and PSpice\cite{pspice}. Applying MNA to a given circuit leads to systems of differential-algebraic equations (DAEs), which can generally be written as systems of implicit differential equations \cite{denk1991}
\begin{align}
    \mb{F}(\mb{x}', \mb{x}, t, \mb{p}) = \mb{0}, \quad \mb{x}(0) = \mb{x}_0, \label{eq:int_dae}
\end{align}
where the Jacobian $\mb{J}_{\mb{x}'}(\mb{F})$, of $\mb{F}$ w.r.t.~$\mb{x}'$, is singular and $\mb{p}$ are the design or uncertainty parameters. Intuitively, one can think of DAEs as ordinary differential equations (ODEs) that are constrained to a manifold determined by (hidden) constraints on the solution variables $\mb{x}$. We aim to exploit the special structure of the DAEs arising from MNA by using the dissection index\cite{jansen2014} to propose an approach for learning electrical circuits more accurately and efficiently. More concretely, we use the dissection index to decouple the DAEs into sets of ODEs and purely algebraic equations, such that the entire dynamics of the solution may then be found only using the ODEs, while the algebraic equations may be used to recover the entire solution.

In the following, \autoref{sec:md} introduces MNA and DAEs in more detail and states some well-known results. Afterwards, \autoref{sec:di} outlines the dissection index and showcases its properties using example circuits. The new approach is then presented on an abstract level in \autoref{sec:ial}, and on a numerical level in \autoref{sec:ne}. Preliminary conclusions about the effectiveness of the approach and future research directions are given in \autoref{sec:cfr}.

\section{MNA and DAEs}
\label{sec:md}
We first look at the system of DAEs that results from MNA when not considering controlled sources. As they are of crucial importance for engineering applications however, we will note whether extensions including controlled sources are available or missing at the appropriate times. Borrowing the notation from Tischendorf\cite{tischendorf1999}, the system of MNA reads
\begin{subequations}
    \label{eq:md_mnai}
    \begin{align}
            \A{C} \ddt \qC \big( \AT{C} \vphi \big) + \A{R} \gR \big( \AT{R} \vphi \big) + \A{L} \i{L} + \A{V} \i{V} + \A{I} \i{s}(t) &= \mb{0}\\ 
            \ddt \phiL \big( \i{L} \big) - \AT{L} \vphi &= \mb{0}\\
            \AT{V} \vphi - \v{s}(t)&= \mb{0},
    \end{align}
\end{subequations}
where $\gR$, $\phiL$ and $\qC$ represent functions modeling resistive, capacitive or inductive devices respectively. The terms for independent current and voltage sources are given by $\i{s}(t)$ and $\v{s}(t)$, while $\vphi$ denotes the vector of nodal potentials and $\i{L} $, $\i{V}$ are the currents flowing through branches containing inductors or voltage sources respectively. The last ingredient is given by the incidence matrices $\A{*}$, where $*$ indicates the device type. These collect the branch to node relations of the underlying electrical network, when considering the branches and nodes as edges and vertices of a directed graph. In the context of design optimization or uncertainty quantification, the device functions $\gR$, $\phiL$ or $\qC$ could e.g.~depend on (subsets of) the parameters $\mb{p}$.

\textcolor{red}{In order to obtain a version of \eqref{eq:md_mnai} that is better suited to analysis and implementation, we consider the device function Jacobians
\begin{align}
    \mb{G} \big( \AT{R} \vphi \big) \coloneqq \mb{J}_{\AT{R} \vphi}(\gR), \quad \mb{L}(\i{L}) \coloneqq \mb{J}_{\i{L}}(\phiL), \quad \mb{C} \big( \AT{C} \vphi \big) \coloneqq \mb{J}_{\AT{C} \vphi}(\qC), \label{eq:md_dfj}
\end{align}
where we use the same notation for the Jacobians as in \autoref{sec:int}.} Inserting \eqref{eq:md_dfj} into the original system, while linearizing $\gR$ and writing everything in matrix form, yields
\begin{align}
    \begin{bmatrix}
        \A{C}^{\phantom{\T}} \mb{C} \big( \AT{C} \vphi \big) \AT{C} & &\\
        & \mb{L}(\i{L}) &\\
        & & \mb{0}
    \end{bmatrix} \ddt \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{R}^{\phantom{\T}} \mb{G} \big( \AT{R} \vphi \big) \AT{R} & \A{L} & \A{V}\\
        -\AT{L} & &\\
        -\AT{V} & &
    \end{bmatrix} \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{I} \i{s}(t)\\
        \mb{0}\\
        \v{s}(t)
    \end{bmatrix} = \mb{0}, \label{eq:md_mnaii}
\end{align}
assuming that $\phiL$ and $\qC$ do not explicitly depend on time.

\textcolor{red}{We note that \eqref{eq:md_mnaii} readily extends to multiport devices\cite{tischendorf1999}. A general inductive $n$-port for example, can also be modeled by a function $\phiL(\i{L}) = \big[ \phi_1(\i{L}), \dotsc, \phi_{n-1}(\i{L}) \big]^\T$, however the Jacobian is not necessarily diagonal in this case as the component functions of $\phiL$ can each depend on all the currents $\i{L} = [i_1, \cdots, i_{n-1}]^\T$. Regarding the incidence matrices, one chooses a reference node $\varphi_0$ for the multiport device and then considers $n-1$ branches from the remaining $n-1$ nodes to the reference. Interpreting the multiport as a single node and using Kirchhoff's current law then gives $i_0 = -\sum_{k=1}^{n-1} i_k$ for the reference current, when orienting all currents to point toward the device. Following this approach, there is then no difference in treating multiports compared to one-ports\cite{tischendorf1999}.}

\subsection{Index concepts}
\label{subsec:ic}
\textcolor{red}{Before outlining the dissection index, we want to give a brief introduction to index concepts more generally. There is no unique definition of the index of a DAE, but rather there are multiple related concepts that each possess different strengths and weaknesses\cite{mehrmann2015}. One important aspect that unites these ideas is that they agree in key cases, e.g.~when looking at linear DAEs, and the same also holds true for the dissection index. To emphasize the practical importance of the notion of index, we take a closer look at the perturbation index. It is based on a perturbed version of the DAE \eqref{eq:int_dae} (we leave out the parameters $\mb{p}$ for conciseness)
\begin{align}
    \mb{F}(\mbh{x}', \mbh{x}, t) = \boldsymbol{\varepsilon}(\mbh{x}', \mbh{x}, t), \quad \mbh{x}(0) = \mbh{x}_0, \label{eq:ic_pdae}
\end{align}
where $\varepsilon$ is a sufficiently smooth perturbation, such that the required derivatives exist.
\begin{definition}
    \label{def:ic_pi}
    Let $\mb{x}$ be a solution of the unperturbed DAE, then the DAE is said to have perturbation index $\nu \in \mathbb{N}$, if $\nu$ is the smallest natural number such that, for any sufficiently smooth solution $\mbh{x}$ of \eqref{eq:ic_pdae}, there exists $c \in \mathbb{R}$ with
    \begin{align*}
        \| \mbh{x} - \mb{x} \| \leq c \big( \| \mbh{x}_0 - \mb{x}_0 \| + \| \boldsymbol{\varepsilon} \|_\infty + \| \boldsymbol{\varepsilon}' \|_\infty + \dotsb + \| \boldsymbol{\varepsilon}^{(\nu - 1)} \|_\infty \big)
    \end{align*}
    for an appropriate norm $\| \cdot \|$ and the right hand side small enough.
\end{definition}
The idea behind this definition is to capture the impact of perturbations on the solution, as the name suggests. Usually these pertubations are assumed small, in the sense that $\| \boldsymbol{\varepsilon} \|_\infty \ll 1$, but fast changing, such that $\| \boldsymbol{\varepsilon}^{(\nu)} \|_\infty$ may grow very quickly in $\nu$.}

\paragraph{Example}
\begin{figure}[b]
    \begin{center}
        \includegraphics[width=0.35\textwidth]{vcr}
    \end{center}
    \caption{Small example circuit for illustrating the perturbation index.}
    \label{fig:ic_vcr}
\end{figure}
\textcolor{red}{To illustrate the perturbation index, and also to hint at its relevance for circuit simulation, we consider the small example given in \autoref{fig:ic_vcr}. Applying MNA to the circuit and introducing a perturbation yields the following DAE
\begin{subequations}
    \label{eq:ic_vcr}
    \begin{align}
        C \ddt \hat{\varphi}_1 + \frac{1}{R} \hat{\varphi}_1 + \hat{i}_\mr{V} &= \varepsilon_1(\mbh{x}', \mbh{x}, t) \label{eq:ic_vcra}\\
        -\hat{\varphi}_1 + v_\mr{s}(t) &= \varepsilon_2(\mbh{x}', \mbh{x}, t), \label{eq:ic_vcrb}
    \end{align}
\end{subequations}
where $\mbh{x} = [\hat{\varphi}_1, \hat{i}_\mr{V}]^\T$. Using \eqref{eq:ic_vcrb} we find
\begin{align}
    \hat{\varphi}_1 = v_\mr{s}(t) - \varepsilon_2(\mbh{x}', \mbh{x}, t). \label{eq:ic_vcrphi}
\end{align}
Inserting \eqref{eq:ic_vcrphi} into \eqref{eq:ic_vcra} and rearranging then gives
\begin{align*}
    \hat{i}_\mr{V} &= \varepsilon_1(\mbh{x}', \mbh{x}, t) - C \ddt \big( v_\mr{s}(t) - \varepsilon_2(\mbh{x}', \mbh{x}, t) \big) - \frac{1}{R} \big( v_\mr{s}(t) - \varepsilon_2(\mbh{x}', \mbh{x}, t) \big).
\end{align*}
Noting that the solution $\mb{x}$ to the unperturbed problem follows directly from the perturbed solution by setting $\boldsymbol{\varepsilon} = \mb{0}$, we obtain
\begin{align}
    \| \mbh{x} - \mb{x} \| = \begin{Vmatrix}
        -\varepsilon_2(\mbh{x}', \mbh{x}, t)\\
        \varepsilon_1(\mbh{x}', \mbh{x}, t) + C \ddt \varepsilon_2(\mbh{x}', \mbh{x}, t) + \frac{1}{R} \varepsilon_2(\mbh{x}', \mbh{x}, t)
    \end{Vmatrix}, \label{eq:ic_vcrpi}
\end{align}
so the unperturbed system corresponding to \eqref{eq:ic_vcr} has perturbation index $\nu = 2$ as \eqref{eq:ic_vcrpi} depends on the first derivative of $\varepsilon_2$.}

\paragraph{Index of MNA}
The structure and index of MNA are well understood when only considering independent sources\cite{tischendorf1999} (as in \eqref{eq:md_mnaii}), but also when including controlled sources\cite{estevez2000ijcta}. For the case without controlled sources, there exists the following well-known topological index result.
\begin{theorem}
    \label{th:ic_imna}
    Assuming the matrix-valued functions $\mb{G} \big( \AT{R} \vphi \big)$, $\mb{L}(\i{L})$, $\mb{C} \big( \AT{C} \vphi \big)$ from \eqref{eq:md_dfj} positive definite:
    \begin{enumerate}
        \item The index of \eqref{eq:md_mnaii} is $\nu \leq 2$.
        \item The index of \eqref{eq:md_mnaii} is $\nu \leq 1$, if and only if there are no loops consisting only of capacitors and voltage sources containing at least one voltage source and no cutsets consisting only of inductors and current sources.
    \end{enumerate}
\end{theorem}
The result can be found in terms of the tractability index\cite{tischendorf1999}, the differentiation index\cite{estevez2000ijcta} and the dissection index\cite{jansen2014}. We again note that there also exist extensive results about when the index of MNA including controlled sources does not exceed $\nu = 2$\cite{estevez2000ijcta}.

\section{Dissection index}
\label{sec:di}
\textcolor{red}{We now focus on the dissection index. As mentioned in \autoref{sec:int}, the key property of this index concept is that it enables the decoupling of a DAE into an ODE and a set of purely algebraic equations. This is clearly different from the perturbation index, however other index concepts such as the tractability and differentiation indices also make such a decoupling possible. Still, the dissection index maintains some advantages over these concepts, as it provides a simple algorithmic procedure for the decoupling that is similar to the tractability index, and it poses less strict smoothness assumptions. In the case of MNA without controlled sources \eqref{eq:md_mnaii}, it is even possible to find a purely topological decoupling based on the dissection index\cite{jansen2014}. We will not make use of this topological decoupling here however, but rather consider the dissection index for a more general class of DAEs, to formulate the assumptions that are necessary for our method to work also for DAEs other than \eqref{eq:md_mnaii}.}

\textcolor{red}{We consider a DAE in standard form\cite{jansen2014},
\begin{align}
    \mb{M}(\mb{x}) \ddt \mb{x} + \mb{K}(\mb{x}) \mb{x} + \mb{f}(t) = \mb{0}, \quad \ker \mb{M} \supsetneq \{ \mb{0} \}, \label{eq:di_dae}
\end{align}
where the matrix-valued functions $\mb{M}$ and $\mb{K}$ are derived from \eqref{eq:int_dae} by defining
\begin{align*}
    \mb{M}(\mb{x}) = \mb{J}_{\mb{x}'} \big( \mb{F}(\mb{x}', \mb{x}, t) \big), \quad \mb{K}(\mb{x}) = \mb{J}_{\mb{x}} \big( \mb{F}(\mb{x}', \mb{x}, t) \big).
\end{align*}
Note that the description of MNA in \eqref{eq:md_mnaii} is precisely of this form. In the following, we will demonstrate the first two steps of the dissection index when applied to systems of the form of \eqref{eq:di_dae}, while stating the assumptions of our approach. Appendix \autoref{sec:dim} contains additional remarks showing that these assumptions are fulfilled by \eqref{eq:md_mnaii}.}

### 2024-01-22 doc/fig/ial.tex
revision

% landscape
\begin{tikzpicture}
    \node[rectangle, rounded corners, draw=black, semithick] (dae) at (0,0) {DAE};
    \node[rectangle, rounded corners, draw=black, semithick] (ode) at (2,-2) {\textcolor{TUDa-1a}{ODE}};
    \node[rectangle, rounded corners, draw=black, semithick] (ae) at (2,2) {\textcolor{TUDa-9a}{AE}};
    \node[rectangle, rounded corners, draw=black, semithick] (dd) at (5.5,-2) {\textcolor{TUDa-1a}{Differential DOFs}};
    \node[rectangle, rounded corners, draw=black, semithick] (ad) at (9.5,2) {\textcolor{TUDa-9a}{Algebraic DOFs}};

>>>>

% landscape
\begin{tikzpicture}
    \node[rectangle, rounded corners, draw=black, semithick] (dae) at (0,0) {DAE};
    \node[rectangle, rounded corners, draw=black, semithick] (ode) at (2,-2) {\textcolor{TUDa-1a}{ODE}};
    \node[rectangle, rounded corners, draw=black, semithick] (ae) at (2,2) {\textcolor{TUDa-9a}{AE}};
    \node[rectangle, rounded corners, draw=black, semithick] (dd) at (5.5,-2) {\textcolor{TUDa-1a}{Differential variables}};
    \node[rectangle, rounded corners, draw=black, semithick] (ad) at (9.5,2) {\textcolor{TUDa-9a}{Algebraic variables}};
